Adaptive Waveform Inversion for Transmitted Wavefields

### 2024-01-21 repro/papers/awi/paper.tex
final maybe

\title{Notes on AWI}
\author{William W. Symes}

\begin{abstract}
goes here.
\end{abstract}

\section{Introduction}

\section{AWI and why it works}

The version of AWI introduced by Michael Warner and Lluis Guasch \cite[]{Warner:16,GuaschWarnerRavaut:GEO19} rests on the assumption that seismic waves are governed by linear acoustics, and that each shot is associated to an isotropic point source with known location $\bx_s$ and wavelet $w(t;\bx_s)$. That is, the pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the shot location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
 $$
 \frac{\partial p}{\partial t} = - \kappa \nabla \cdot {\bf v} +
w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s);
\frac{\partial {\bf v}}{\partial t} = - \beta \nabla p; 
p, {\bf v} = 0 \mbox{ for }  t \ll 0.
$$
For sake of brevity, define the model vector $m=(\kappa,\rho)$. The forward map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf x}_s)\}$, in which shot and receiver positions ${\bf x}_s, {\bf x}_r$ define the acquisition geometry.

Acoustic inversion, for present purposes, means: given time interval $[0,T]$, source and receiver locations $(\bx_s,\bx_r)$, source wavelet $w_* \in C_0^{\infty}(\bR)$ and data traces $d(\bx_r,\cdot;\bx_s) \in L^2([0,T])$, find a model $m$ so that $F[m]w_* \approx d$. The simplest version of FWI concretizes this task by asking for a model $m$ minimizing the mean square error
\begin{equation}
  \label{eqn:fwi}
  J_{\rm FWI}[m;d] = \frac{1}{2}\|F[m]w_*-d\|^2
\end{equation}
Practical versions of FWI include various types of regularization and homotopy, see the sources cited in the Introduction. As mentioned there, local iterative optimization methods tend to stagnate at highly sub-optimal (``cycle-skipped'') estimates of $m$ unless the initial estimate is of quite high quality, often available only at considerable cost in effort and data collection.
%For now, assume that the data $d({\bf x}_r,t;{\bf x}_s)$ is the output of the modeling operator for "true" model $m_*$, that is, the "true" bulk modulus, buoyancy, and wavelet $\kappa_*, \beta_*, w_*$: that is, $d = F[m_*]w_*$.

AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
The extended modeling operator ${\bar F}$ maps extended sources $\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure field. That is, the extended source depends on the receiver location as well as the source location - so there is one acoustic system for each source *and* receiver position - a lot of wave equations! If all of the wavelets for each source are the same, that is, ${\bar w}({\bf x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of $F$. This is the *source-recever* extension, in the terminology of Huang et al. 2015.

AWI assumes that the extended sources are time convolutions of the (known) exact source with an {\em adaptive kernel} $u({\bf x}_r,t;{\bf x}_s)$: that is, $\bar{w} = u * w_*$ - the asterisk denotes convolution in time. Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
$$
\bar{F}[m]\bar{w} = \bar{F}[m](u* w_*) = (F[m]w_*)*u
$$
For notational convenience, introduce the notation $S[m,w_*]$ for the convolution operator with kernel $F[m]w_*$:
\begin{equation}
  \label{eqn:defs}
  S[m,w_*]u = (F[m]w_*)*u
\end{equation}
Warner and Guasch assume (implicitly) that there is always a (near-)zero-residual solution of the extended inversion problem. That is, they assume that for any $m$, there is an extended source $\bar{w}$, or equivalently an adaptive kernel $u_0$ with $\bar{w} = w_**u_0$ for which $\bar{F}[m]\bar{w} = S[m,w_*]u_0\approx d$. If $\kappa \approx \kappa_*, \beta \approx \beta_*$, then $\bar{w}$ should be $\approx w_*$, so the adaptive kernel $u_0$ should be approximately $\delta(t)$ and independent of ${\bf x}_s,{\bf x}_r$. Such a kernel $u_0$ is (approximately) in the null space of multiplication by $t$.

This observation leads to the definition of the Warner-Guasch AWI objective function $\tJz$:

\begin{itemize}
\item[1. ]Given $m$, solve the deconvolution problem $(F[m]w_*)*u_0 = d$ for $u_0$, if necessary in a least squares sense and with Tihonov regularization;. 
\item[3. ]Compute the objective 
  \begin{equation}
    \label{eqn:tJz}
    \tJz[m] = \sum_{x_s,x_r} \left(\frac{\int dt |tu_0|^2}{\int dt |u_0|^2}\right)
  \end{equation}
and its gradient, then update $\kappa,\beta$ using a gradient descent method to minimize $\tJz$.
\end{itemize}

The normalization of the integrand per source and receiver by the $L^2$ norm of $u_0$ is quite important, as will be explained shortly.

It may not be obvious at first glance why this approach to estimating $m$ should be any more effective than FWI. To see why AWI works, in at least some settings, we assume that the spatial dimension is three, and make use of the geometric optics approximation to the solution of the wave equation, in a form due to Hadamard. This approximation is {\em local}: it applies when sources and receivers are sufficiently close. We will discuss its global failure and the consequences for AWI in a later section.

Transient acoustic sources have no DC (zero frequency) component and are relatively small at low frequencies. One was to enforce these characteristics is to assume that the wavelet $w_*$ is the time derivative of a $v_* \in C_0^{\infty}(\bR)$. Set $\lambda(w_*) = \|v_*\|/\|w_*\|$ ($L^2$ norms). $\lambda$ is a proxy for wavelength: it is the reciprocal of the RMS frequency.

For a particular source-receiver pair ${\bf x}_s,{\bf x}_r$ with $\bx_r$ "not too far" from $\bx_s$,, suppressed from the notation,
\begin{equation}
  \label{eqn:hadamard}
  F[m]w_*(t) = a[m]w_*(t-\tau[m]) + Rw_*(t)
\end{equation}
where $R(\bx_r,\bx_s)$ is a bounded operator on $L^2((-\infty,T])$ bounded by a multiple of $\lambda(w_*)$. This ``single arrival'' approximation  follows from the expansion of the fundamental solution of the wave equation due to Hadamard, see for instance \cite{Friedlander:75}. This construction is sketched in Appendix B. Note that in general the approximation \ref{eqn:hadamard} is only local: for smooth $m$, it holds for $\bx_r$ sufficiently close to $\bx_s$, but fails at larger distances.

For approximations with $O(\lambda(w_*)$ remainders, write
\begin{equation}
  \label{eqn:pwe}
  F[m]w_*(t) \approx a[m]w_*(t-\tau[m])
\end{equation}
Thanks to the approximation \ref{eqn:pwe},
\begin{equation}
  \label{eqn:spwe}
S[m,w_*]u(t) \approx a[m](w_**u)(t-\tau[m])
\end{equation}
Assume that the data is noise-free:
$$
d(t)=F[m_*]w_*(t) \approx a[m_*]w_*(t-\tau[m_*])
$$
The (approximate) solution of $S[m,w_*]u_0=d$ is obvious by inspection: $u_0(t) = a[m_*]/a[m]\delta(t-\tau[m_*]+\tau[m])$. There is immediately a big problem: this $u_0$ is not square-integrable, and the AWI objective $\tJz$ is undefined. To fix this, it's necessary to treat $u_0$ as the solution of a regularized least squares problem: define
$$
u_0^{\epsilon} = \mbox{argmin}_u (\|S[m,w_*]u-d\|^2 + \epsilon^2\|u\|^2)
$$
$$
\approx \mbox{argmin}_u (\|a[m](w_**u)(\cdot-\tau[m])-a[m_*]w_*(\cdot-\tau[m_*])\|^2 + \epsilon^2\|u\|^2)
$$
Use the temporary abbreviations $a_*=a[m_*], a=a[m], \tau_*=\tau[m_*], \tau=\tau[m]$, and write in terms of the Fourier transforms $\hat{u}_{\epsilon},\hat{w_*}$ of $u_0^{\epsilon}$ and $w_*$:
$$
\hat{u}_{\epsilon} \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w_*}\hat{u}e^{i\omega \tau}-a_*\hat{w_*}e^{i\omega\tau_*}|^2 + \epsilon^2 |\hat{u}|^2\right)
$$
The normal equation is
$$
(a^2 |\hat{w}_*|^2 +\epsilon^2)\hat{u} = a a_*|\hat{w}_*|^2e^{i\omega(\tau_*-\tau)}
$$
the solution of which is
$$
\hat{u}_{\epsilon} \approx \frac{a^*}{a}\hat{g}_{\frac{a^*}{a}\epsilon} e^{i\omega(\tau_*-\tau)}
$$
where
$$ 
\hat{g}_{\epsilon} = \frac{a_*^2|\hat{w}_*|^2}{a_*^2|\hat{w}_*|^2 + \epsilon^2}.
$$
Since $w_* \in C_0^{\infty}$, $\hat{w_*}(\omega) \rightarrow 0$ faster than any negative power of $\omega$ as $|\omega| \rightarrow \infty$ and is analytic, whence $\hat{g}_{\epsilon}: \epsilon > 0$ tends to $1$ almost everywhere as $\epsilon \rightarrow 0$. Hence the inverse Fourier transform $g_{\epsilon}$ tends to $\delta$ in the sense of distributions, and 
$$
u_0^{\epsilon}(t) \approx \frac{a^*}{a}g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))
$$
to a multiple of a shifted $\delta$, in fact exactly the distribution solution of $\bar{S}[m,w_*]u=d$. However for $\epsilon>0$, $u_0^{\epsilon}$ is square integrable.

\cite{HuangSymes:SEG15a} presented an analysis of the unweighted AWI objective (numerator of the right hand side in the definition \ref{eqn:tJz}), based in turn on ideas presented in Song's thesis \cite[]{Song:94c}. With $u$ replaced by $u_0^{\epsilon}$,  this unweighted objective is
$$
\int dt |tu_0^{\epsilon}|^2 \approx \frac{a_*^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\epsilon}(t-(\tau_*-\tau))|^2
$$
$$
=\frac{a_*^2}{a^2} \int dt\, (t+(\tau_*-\tau))^2|g_{\frac{a^*}{a}\epsilon}(t)|^2
$$
The linear (in t) term vanishes, since $g$ is necessarily even in time (since it Fourier transform is real). So this is
$$
=\frac{a_*^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\epsilon}(t)|^2
+(\tau_*-\tau)^2\frac{a_*^2}{a^2}\|g_{\frac{a^*}{a}\epsilon}\|^2
$$
The first term tends to zero as $\epsilon \rightarrow 0$, since $g_{\epsilon}$ is a delta-family, so you can make it negligible with proper choice of $\epsilon$. The second term suggests that this function is related to the travel-time error $\tau_*-\tau$.

The relation is obscured by the presence of of the amplitude ratio $a_*/a$ both as a multiplier and in the scaling of $\epsilon$. Since the amplitudes depend on $m$, their presence may influence the location (or even the presence) of  stationary points of $\tJz$ for models differing in their travel time predictions. So stationary points of the unweighted AWI objective may not yield slowness models that match travel-times with the data, even in the noise-free case (even though the examples reported in Huang and Symes SEG 2015, and other works cited there, suggest that sometimes this is the case).

The scaling by trace norms of $u_0^{\epsilon}$ in the definition \ref{eqn:tJz} eliminates the possibility of non-travel-time stationary points. Since the objective function is built up trace-by-trace, use the same approximation for the norm of $u_0^{\epsilon}$: 
$$
\|u_0^{\epsilon}\| \approx \frac{a^*}{a}\|g_{\frac{a^*}{a}\epsilon}\|
$$
So
$$
J_0[m,w_*] \approx \sum_{{\bf x}_s,{\bf x}_r} (\tau[m_*]({\bf x}_s,{\bf x}_r)-\tau[m]({\bf x}_s,{\bf x}_r))^2
$$
ignoring the $\|tg_{...}\|^2$ term as being negligible for small enough $\epsilon$.

In other words, the AWI objective is *exactly* an approximation to the travel-time tomography objective. And that's why it works - when geometric optics in the form given here is an accurate approximation to the Green's function.

It remains to see what geometric optics has to say about the penalty objective $\tilde{J}_{\alpha}[m,w_*]$, how geometric optics might fail, and what happens to AWI when it does.

\section{A Penalty Version of AWI}
The adaptive kernel $u_0^{\epsilon}$ appearing in the definition \ref{eqn:tJz} is the solution of a regularized least squares problem. This observation suggests augmenting the regularized least squares objective by a penalty proportional to the right-hand side of \ref{eqn:tJz}:
\begin{equation}
  \label{eqn:Ja}
  \Ja[m,u;d] = \frac{1}{2}\left(\|S[m,w_*]u-d\|_d^2 + \epsilon^2\|u\| + \alpha^2\|tu\|_m\right).
\end{equation}
Here $\|u\|^2 = \langle u, u \rangle$ is the (unweighted) $L^2$ norm, and $\|\cdot\|_m$ is an $(m,d)$-dependent (weighted) norm: with
\[
  u_0^{\epsilon}[m,d] = \mbox{arg min}_u \Jz[m,u;d],
\]
define
\[
  Wu (\bx_r,t;\bx_s) = \sum_{\bx_r,\bx_s}\frac{1}
  {\int\,ds\,|u_0^{\epsilon}[m,d](\bx_r,s;\bx_s)|^2}u(\bx_r,t;\bx_s)
\]
and
\begin{equation}
  \label{eqn:defmnorm}
  \|u\|^2_m = \langle u, W u \rangle.
\end{equation}
With this notation,
\[
  \tJz[m;d] = \frac{1}{2}\|tu^{\epsilon}_0[m;d]\|^2_m
\]
In the definition of $\tJz$, a quadratic is evaluated at the minimizer of $\Jz[m,u;d]$. This suggests the possibility of evaluating the quadratic objective $\Ja[m,u;d]$ at its own minimizer:
\begin{equation}
  \label{eqn:tJa}
  \tJa[m;d] = \min_u \Ja[m,u,d] = \Ja[m,\ua[m;d];d]
\end{equation}
This is the so-called variable projection reduction of the quadratic $\Ja$.
Note that the right-hand side of the definition of $\tJz$ is the last term on the right-hand side of $\Ja$, scaled by $\alpha^{-2}$ and evaluated at the minimizer of the sum of the first two terms, that is, $\Ja$ for $\alpha=0$.

\begin{theorem}
  \label{thm:basic}
Suppose that $S \in {\cal L}(U,D)$, the is a bounded linear maps from (Hilbert space) domain $U$ (``model space'') with inner product $\langle \cdot, \cdot \rangle_m$ to range $D$ with inner product $\langle \cdot, \cdot \rangle_d$. Assume that $S$ is coercive, that is, there exists $\epsilon >0$ so that $\|Su\|_d \ge \epsilon\|u\|_m$ for $u \in U$.
Suppose also that $A \in {\cal L}(U,U)$, and that $d \in D$. Define the function $J_{\alpha}$ on $U$ by 
\begin{equation}
  \label{eqn:eq1}
  J_{\alpha}(u) = \frac{1}{2}(\|Su-d\|^2_d + \alpha^2\|Au\|_m^2).
\end{equation}
Then for $\alpha \ge 0$, $J_{\alpha}$ possesses a unique minimizer $u_{\alpha} \in U$, for which
\begin{equation}
  \label{eqn:eq0}
  \tilde{J}_{\alpha} = \min_{u \in U} J_{\alpha}(u) = J_{\alpha}(u_{\alpha}),
\end{equation}
and
\begin{equation}
  \label{eqn:eq4}
  \lim_{\alpha \rightarrow 0} \frac{1}{\alpha^2}  (\tilde{J}_{\alpha}-\tilde{J}_0).= \frac{1}{2}\|Au_0\|^2
\end{equation}
This limit is uniform over bounded sets of $(S, A) \in {\cal L}(U,D) \times {\cal L}(U, U)$.
\end{theorem}

For completeness, the proof is given in an appendix.

This relation will be applied in the context of a model space $U$ possessing a reference Hilbert norm $\|\cdot\|$ and a family of equivalent norms $\|\cdot\|_m$ depending on a parameter $m \in M$, in which $M$ is an open subset of a Banach space $X$. The dependence is through a an operator-valued function $W$ on $M$, taking values in cone ${\cal B}(U)$ of bounded self-adjoint positive-definite operators on $U$. On the other hand, we assume that the norm on $D$ is fixed, independent of $M$, and treat it as a reference norm.

>>>>

\title{Adaptive Waveform Inversion for Transmitted Wavefields}
\author{William W. Symes}

\begin{abstract}
Under some restricted circumstances, Adaptive Waveform Inversion
applied to transmitted wave data yields estimates of index of
refraction (or wave velocity) similar to those obtained by travel time inversion.
\end{abstract}

\section{Introduction}
Full waveform inversion (FWI) refers to the inference of mechanical
parameter distribution within a material object  (the earth, a human
body, a manufactured artifact,...) from the measurement of wave motion
at its surface. Over the last several decades, this task has come to
be carried out by minimizing the difference between computationa
prediction of wave measurement versus actual observed
measurements. The mechanical parameters to be inferred appear as
coefficients in systems of partial differential equaitons governing
the wave motion, and discrete representations of these parameters form
the unknown vectors in the optimization problems thus generated.

While widely used in recent years, FWI exhibits a pathology when wave
velocities or their parametric equivalents are amongst the parameters
to be estimated: the excessive sensitivity of predicted waveforms to
changes in velocity leads to oscillations in the difference measures,
especially for the most commonly used mean-square measure. Local
optimization algorithms (relatives of Newton;'s method) are the only practical
methods to generate sequences of approximate optimal
parameters. However the oscillations in the difference measures lead
these sequences to become entrapped at near parameter distributions
far from the optimum, and thus do not extract the maximum information
from the data \cite[]{VirieuxOperto:09}.

Various modifications of straighforward best-fit FWI have been
suggested to overcome this ``cycle-skipping'' pathology. One of these
is Adaptive Waveform Inversion \cite[]{Warner:16} (AWI). This
technique has been successfully applied to industry-scale seismic data
\cite[]{GuaschWarnerRavaut:GEO19,Warneretal:SEG21} as well as to
ultrasonic imaging of the the human brain
\cite[]{Guaschetal:NPJDM20}.

While heuristic justifications for AWI (and
many details of implementation) are discussed in these and other
works, a clear explanation for its effectiveness is 
so far missing from the literature. In this short note, I show that
for active sensing of acoustic materials with unique arrival times of
transmitted waves
(that is unique rays of geometric asymptotics between sources and
receivers), the AWI objective function is an approximation to the
mean-square time-of-travel misfit. Thus AWI and travel time fitting
lead to similar material parameter estimates. Since the latter is
widely believed to be a near-optimal method for wave velocity
estimation, this demonstration explains roughly ``why it works''.

The reader will
note that a normalization component of the AWI objective function
definition is very
important in reaching this conclusion. The connection between
objective normalization and the close relation to travel time
inversion is completely independent from 
the reasons for normalizing the objective advanced in the
literature \cite[]{Warner:16,LiAlkhalifah:21}.

After establishing
the class of problems to be studied and notations for their
description in the following section, I present the asymptotic
analysis leading to the main conclusion in the next. The close relation
between AWI objective and travel time misfit no longer mecessarily
holds if multiple rays connect source and receiver. I discuss these
and other aspects of AWI and related approaches in a brief final section.

\section{Adaptive Waveform Inversion}

The version of AWI introduced by \cite{Warner:16} is based on an
active source acoustic model of data generation. Each acoustic source
is modeled an isotropic point radiator with known location $\bx_s$ and
wave form (``wavelet'') $w(t;\bx_s)$. Similarly, receivers are modeled
as isotropic point sensors, recording pressure at points ${\bf x}_r$. The pressure and velocity fields $p({\bf x},t;{\bf x}_s)$, ${\bf v}({\bf x},t;{\bf x}_s)$ for the source location ${\bf x}_s$ depend on the bulk modulus $\kappa({\bf x})$, buoyancy $\beta({\bf x})$ (reciprocal of the density $\rho({\bf x})$), and wavelet $w(t;{\bf x}_s)$ through the acoustic system
\begin{eqnarray}
  \label{eqn:awe}
 \frac{\partial p}{\partial t} & = &- \kappa \nabla \cdot {\bf v} +
w(t;{\bf x}_s) \delta({\bf x}-{\bf x}_s); \nonumber \\
\frac{\partial {\bf v}}{\partial t} & = & - \beta \nabla p; \\ 
p, {\bf v} & = & 0 \mbox{ for }  t \ll 0.
\end{eqnarray}
For sake of brevity, define the model vector $m=(\kappa,\rho)$. The forward map or {\em modeling operator} is $F[m]w = \{p({\bf x}_r,t;{\bf x}_s)\}$, in which source and receiver positions ${\bf x}_s, {\bf x}_r$ define the acquisition geometry.

Acoustic inversion, for present purposes, means: given recording time interval
$[0,T]$, source and receiver locations $(\bx_s,\bx_r)$, source wavelet
$w(t;\bx_s)$ and data traces $d(\bx_r,\cdot;\bx_s)$, find a model $m$
so that $F[m]w \approx d$. The
simplest version of FWI concretizes this task by asking for a model
$m$ minimizing the mean square error
\begin{equation}
  \label{eqn:fwi}
  J_{\rm FWI}[m;d] =\frac{1}{2} \sum_{{\bf x}_r,{\bf x}_s} \int\,dt\, |(F[m]w)({\bf x}_r,t;{\bf x}_s) - d({\bf x}_r,t;{\bf x}_s)|^2
\end{equation}
It is convenient to introduce the $L^2$ norm $\| \cdot \|$ as
shorthand for the right-hand side of the definition \ref{eqn:fwi}:
\[
  J_{\rm FWI}[m;d]= \frac{1}{2}\|F[m]w-d\|^2.
\]
In the sequel I will use the $L^2$ norm notation to stand for
mean-square over whatever variables make sense.

Note that the wavelet $w$ is assumed known in this formulation of the
FWI problem. In fact, even if actual acoustic sources are
well-approximated by isotropic point radiators, their wave forms are
usually poorly determined and may reasonably be
included as  parameters to be estimated by inversion of the
data. However for the present discussion we will regard $w$ as known.

Practical versions of FWI include various types of regularization and
imposition of constraints on $m$, see the sources cited in the
Introduction. As mentioned there, local iterative optimization methods
are the only practical approach to minimizing $J_{\rm FWI}$ due to the
computational cost of approximating the forward map $F$ and related
operators, but these tend to stagnate at highly sub-optimal
(``cycle-skipped'') estimates of $m$ unless the initial estimate is of
quite high quality, often available only at considerable cost in data
collection and processing.
%For now, assume that the data $d({\bf x}_r,t;{\bf x}_s)$ is the output of the modeling operator for "true" model $m^*$, that is, the "true" bulk modulus, buoyancy, and wavelet $\kappa^*, \beta^*, w^*$: that is, $d = F[m^*]w^*$.

AWI is based on extending the modeling operator $F$, that is, enlarging its domain.
The extended modeling operator ${\bar F}$ maps extended sources
$\bar{w}({\bf x}_r,t;{\bf x}_s)$ to the same sampling of the pressure
field. That is, the extended source depends on the receiver location
as well as the source location - so there is one acoustic system for
each source {\em and} receiver position - a lot of wave equations! If all
of the wavelets for each source are the same, that is, ${\bar w}({\bf
  x}_r,t;{\bf x}_s) = w(t;{\bf x}_s)$ is independent of receiver
position, then ${\bar F}[m]\bar{w} = F[m]w$. That is, $F$ is a special
case, or restriction, of ${\bar F}$, so ${\bar F}$ is an extension of
$F$. This is the {\em source-receiver} extension, in the terminology of \cite{HuangSymes2015SEG}.

AWI assumes that the extended sources are time convolutions of the
(known) source $w$ with an {\em adaptive filter} $u({\bf
  x}_r,t;{\bf x}_s)$: that is,
\[
  \bar{w}({\bf x}_r,t;{\bf x}_s) = (u * w)({\bf  x}_r,t;{\bf x}_s)
  = (u({\bf  x}_r,\cdot;{\bf x}_s)*w)(t).
\]
The asterisk denotes convolution in time. That is,
$u*w$ denotes the result of convolving each trace in $u$ by $w$. I
will also use the asterisk to denote the tracae-by-trace convolution
of two data sets with the same set of source and receiver coordinates,
that is, the data set consisting of time convolutions of traces with the same source
and receiver coordinates. The meaning will be clear from context.

Since linear acoustics is time-translation invariant, its solution commutes with time convolution, that is,
\[
 \bar{F}[m]\bar{w} = \bar{F}[m](u* w) = (F[m]w)*u
\]
Since the known source $w$ is not adjusted during the inversion, 
introduce the notation $S[m]$ for the convolution operator with kernel $F[m]w$:
\begin{equation}
  \label{eqn:sdef}
  (F[m]w)*u \equiv S[m]u
\end{equation}
\cite{Warner:16} assume (implicitly) that there is always an exact
solution of the extended inversion problem. That is, they assume that
for any $m$, there is an extended source $\bar{w}$, or equivalently an
adaptive filter $u[m;d]$ with $\bar{w}=w*u[m;d]$, for which
$S[m]u[m;d]= d$. [This proposition is dubious, but is easy to work
around, as will be seen below. Also, \cite{Warner:16} also formulate
AWI in terms of the inverse filter, that is, $u$ for which
$u*d = F[m]w$, and in fact recommend that definition. The present
discussion uses only the forward filter.] If
$\kappa = \kappa^*, \beta = \beta^*$, then $\bar{w}$ should be $= w$,
so the adaptive kernel $u[m;d]$ should be $=\delta(t)$ for all
${\bf x}_s,{\bf x}_r$. Such a kernel $u[m;d]$ is in the null space of
multiplication by $t$.

This observation leads to the definition of the AWI objective function $J_{\rm AWI}$:
\begin{equation}
  \label{eqn:tJz}
  J_{\rm AWI}[m;d] = \sum_{\bx_s,\bx_r} \left(\frac{\int dt |tu[m;d](\bx_r,t;\bx_s)|^2}{\int dt |u[m;d](\bx_r,t;\bx_s)|^2}\right)
\end{equation}
in which $u[m;d]$ is as defined above, that is, the solution of $S[m]u[m;d]
= d$.

Adaptive waveform inversion consists in iterative minimization of
$J_{\rm AWI}$: choose an initial estimate $m_0=(\kappa_0,\beta_0)$ of the
model, then update $m=(\kappa,\beta)$ repeatedly using a gradient
descent method. \cite{Warner:16} describes the computation of $J_{\rm AWI}$
and its gradient in detail, and their use in a quasi-Newton
optimization algorithm.

The trace-by-trace normalization in definition \ref{eqn:tJz} of each
trace in $tu[m;d]$ by the
$L^2$ norm of the corresponding trace in $u[m;d]$ is quite important, as will be explained shortly.

\section{Why it works}

It may not be obvious at first glance why this approach to estimating
$m$ should be any more effective than FWI. To see why AWI works, in at
least some settings, we assume that the spatial dimension is three,
and make use of the geometric asymptotics (ray theory) approximation
to the solution of the wave equation. This approximation is {\em
  local}: it applies when sources and receivers are sufficiently
close. We will discuss its global failure and the consequences for AWI
in the final section.

Transient acoustic sources have no DC (zero frequency) component and
are relatively small at low frequencies. One way to enforce these
characteristics is to assume that the wavelet $w$ is the time
derivative of another wavelet $v$, also transient ($v(t;{\bf x}_s)=0$
for $|t| \gg 0$). Set $\lambda(w) = \|v\|/\|w\|$. $\lambda$ is a proxy for wavelength: it is the reciprocal of RMS frequency.

The geometric asymptotics approximation for solution of the point
radiator problem \ref{eqn:awe} expresses the pressure field in terms
of $m$-dependent geometric amplitude $a[m]({\bf x}_s,{\bf x}_r)$ and
travel time $\tau[m]({\bf x}_s,{\bf x}_r)$, and implies that
\begin{equation}
  \label{eqn:hadamard}
  (F[m]w)({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) w(t-\tau[m]({\bf x}_s,{\bf x}_r)) 
\end{equation}
In this approximation, and in other uses of the symbol $\approx$, the
error is $O(\lambda(w))$.

Approximation \ref{eqn:hadamard} follows from a
``progressing wave''  expansion of the fundamental
solution (Green's function) of the wave equation due to Hadamard
\cite[]{Friedlander:75,Qian:JCP24}. For smooth (slowly varying)
$m$, it holds for $\bx_r$ sufficiently close to $\bx_s$ that a unique ray of
geometric acoustics connect $\bx_s$ to $\bx_r$. This ``single
arrival'' condition fails (in
general) at larger distance. The precise region of validity depends on
$m$. For the remainder of this section, its validity is assumed.

Applying the approximation \ref{eqn:hadamard}, obtain
\begin{equation}
  \label{eqn:spwe}
S[m]u({\bf x}_r,t;{\bf x}_s) \approx a[m]({\bf x}_s,{\bf x}_r) (w*u)(t-\tau[m]({\bf x}_s,{\bf x}_r))
\end{equation}
Assume that the data is noise-free:
$$
d({\bf x}_r,t;{\bf x}_s)=(F[m^*]w) ({\bf x}_r,t;{\bf x}_s) \approx a[m^*]({\bf x}_s,{\bf x}_r) w(t-\tau[m^*]({\bf x}_s,{\bf x}_r))
$$
The (approximate) solution of $S[m]u[m;d]=d$ is obvious by inspection:
\begin{equation}
  \label{eqn:notl2}
  u[m;d]({\bf x}_r,t;{\bf x}_s) = \frac{a[m^*]({\bf x}_s,{\bf x}_r)}{a[m]({\bf x}_s,{\bf x}_r)}\delta(t-\tau[m^*]({\bf x}_s,{\bf x}_r)+\tau[m]({\bf x}_s,{\bf x}_r)).
\end{equation}
There is immediately a big problem: this $u[m;d]$ is not square-integrable, and the AWI objective $J_{\rm AWI}$ is undefined. To fix this, it's necessary to re-define $u[m;d]$ as the solution of a regularized least squares problem: define
\begin{equation}
  \label{eqn:reg}
  u[m;d] = \mbox{argmin}_u (\|S[m]u-d\|^2 + \sigma^2\|u\|^2)
\end{equation}
$$
\approx \mbox{argmin}_u (\|a[m](w*u)(\cdot-\tau[m])-a[m^*]w(\cdot-\tau[m^*])\|^2 + \sigma^2\|u\|^2)
$$

Note that in these expressions, the source and receiver
coordinates ${\bf x}_s$ and ${\bf x}_r$ act as either passive parameters
or summation indices. Therefore, to simplify the presentation, drop them from the notation
temporarily. Also,  use the temporary abbreviations $a^*=a[m^*], a=a[m], \tau^*=\tau[m^*],
\tau=\tau[m]$.

With these conventions, re-write equation \ref{eqn:reg} in terms of the Fourier transforms $\hat{u}[m;d],\hat{w}$ of $u[m;d]$ and $w$:
\begin{equation}
  \label{eqn:freqreg}
\hat{u}[m;d] \approx \mbox{argmin}_{\hat{u}}\left(\int d\omega |a\hat{w}\hat{u}e^{i\omega \tau}-a^*\hat{w}e^{i\omega\tau^*}|^2 + \sigma^2 |\hat{u}|^2\right)
\end{equation}
The normal equation for the least squares problem \ref{eqn:freqreg} is
$$
(a^2 |\hat{w}|^2 +\sigma^2)\hat{u} = a a^*|\hat{w}|^2e^{i\omega(\tau^*-\tau)}
$$
the solution of which is
$$
\hat{u}[m;d] \approx \frac{a^*}{a}\hat{g}_{\frac{a^*}{a}\sigma} e^{i\omega(\tau^*-\tau)}
$$
where
$$ 
\hat{g}_{\sigma} = \frac{(a^*)^2|\hat{w}|^2}{(a^*)^2|\hat{w}|^2 + \sigma^2}.
$$
depends only the ``true'' model $m^*$. Transient acoustic sources are effectively bandlimited: $\hat{w}(\omega) \rightarrow 0$ faster than any negative power of $\omega$ as $|\omega| \rightarrow \infty$ and is analytic, whence $\hat{g}_{\sigma}:$ tends to $1$ almost everywhere as $\sigma \rightarrow 0$. Hence the inverse Fourier transform $g_{\sigma}$ tends to $\delta$ in the sense of distributions, and 
$$
u[m;d](t) \approx \frac{a^*}{a}g_{\frac{a^*}{a}\sigma}(t-(\tau^*-\tau))
$$
to a multiple of a shifted $\delta$, in fact exactly the distribution
solution of $\bar{S}[m,w]u=d$ given in equation \ref{eqn:notl2}. However for $\sigma>0$, $u[m;d]$ is square integrable.

\cite{HuangSymes2015SEG} presented an analysis of the Matched Source
Waveform Inversion (MSWI) objective function, which is essentially the
numerator of the right hand side in the definition
\ref{eqn:tJz} of the AWI objective. This analysis was based in turn on ideas presented in Song's thesis
\cite[]{Song:94c}. MSWI deals with the extended source $\bar{w}$, but
the same construction applies to the adptive filter $u$. With $u$ replaced by $u[m;d]$,  this MSWI objective is
\[
J_{\rm MSWI}[m;d] = \int dt |tu[m;d]|^2 \approx \frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\sigma}(t-(\tau^*-\tau))|^2
\]
\[
=\frac{(a^*)^2}{a^2} \int dt\, (t+(\tau^*-\tau))^2|g_{\frac{a^*}{a}\sigma}(t)|^2
\]
\begin{equation}
  \label{eqn:MSWI}
  =\frac{(a^*)^2}{a^2} \int dt\, (t^2 + 2t
  (\tau^*-\tau)+(\tau^*-\tau)^2)|g_{\frac{a^*}{a}\sigma}(t)|^2
\end{equation}
As noted by \cite{HuangSymes2015SEG}, the linear term (in t) on the
right-hand side of \ref{eqn:MSWI} vanishes,
since $g$ is necessarily even in time (since its Fourier transform is
real). So
\begin{equation}
  \label{eqn:MSWIasym}
  J_{\rm MSWI}[m;d] \approx
\frac{(a^*)^2}{a^2} \int dt\, t^2|g_{\frac{a^*}{a}\sigma}(t)|^2
+(\tau^*-\tau)^2\frac{(a^*)^2}{a^2}\|g_{\frac{a^*}{a}\sigma}\|^2
\end{equation}
The first term tends to zero as $\sigma \rightarrow 0$, since
$g_{\sigma}$ is a delta-family, so you can make it negligible with
proper choice of $\sigma$. The second term suggests that this function
is related to the travel-time error $\tau^*-\tau$.
